---
title: "Practical Machine Learning - Project 1"
author: "Nick Allen"
date: "Wednesday, June 18, 2014"
output: pdf_document
---

Load the project data into the workspace. The 'ProjectTemplate' package is used to structure the project. 

```{r, message=FALSE, warning=FALSE}
set.seed (825)
library (ProjectTemplate)
load.project( )
```

The training data has been loaded into the environment.

```{r}
dim (pml.training)
dim (pml.testing)
```


The first 5 fields are not useful predictors.  This includes a row identifier, subject name, etc.  These predictors would not produce a generalizable model.

```{r}
exclude <- 1:5
pml.training <- pml.training[, -exclude, with = FALSE]
dim (pml.training)
```

Many of the predictors have a large number of missing values and so very little variation.  These will not be useful for training and are excluded.

```{r}
nzv <- nearZeroVar (pml.training)
pml.training <- pml.training[, -nzv, with = FALSE]
dim (pml.training)
```

Many of the fields have large numbers of missing values.  Exclude these as they will not be useful.

```{r}

# if a column has more than 5% NAs, exclude it
max.nas <- nrow (pml.training) * 0.05
nas.by.column <- colSums (is.na (pml.training)) 
too.many.nas <- names (nas.by.column [ nas.by.column > max.nas])

# exclude the columns
keep <- setdiff (names (pml.training), too.many.nas)
pml.training <- pml.training[, keep, with = FALSE]
```


The original training data set is split 80/20 into a training set and a validation data set to use for cross-validation

```{r}
train.index <- createDataPartition (pml.training$classe, p = 0.80, list = FALSE)[, 1]
train    <- pml.training [ train.index]
validate <- pml.training [-train.index]
```

Train a tree-based, Gradiant Boosting Machine (GBM) model to predict the outcome 'classe' based on the other 159 predictors.  The GBM model performs feature selection automatically.  This is important since there are a large number of potential predictors.

```{r, results='hide'}
control <- trainControl (method = "cv", number = 3, repeats = 1)
profile <- train (classe ~ ., data = train, method = "gbm", trControl = control)
```

```{r}
profile
```

The model found the following predictors to be the most important for predicting the outcome.

```{r}
plot (varImp (profile, scale = FALSE), top = 20)
```

Details which model parameters were most effective and ultimately selected for the final mode.

```{r}
trellis.par.set (caretTheme())
plot (profile, type = c("g", "o"))
```

Using the trained model, predict the 'classe' for the validation set.  Assuming that the provided training and test sets have similar characteristics, this will provide a fair measure of the model's accuracy.

```{r, results='hide'}
validate [, classe.hat := predict (profile, newdata = .SD)]
```

A confusion matrix identifies how good the predictions were.  The model is achieving a >98% level of accuracy.  The results are similar across all reference classes.

```{r}
confusionMatrix(validate$classe.hat, validate$classe)
```

Using the trained model, make predictions on the test data that needs to be submitted as part of this project.

```{r, results='hide'}
pml.testing [, classe.hat := predict (profile, newdata = .SD)]
answers <- pml.testing [["classe.hat"]]
```

Output the results so they can be submitted.

```{r}
#
# write the solutions for the test set to separate files for submission
#
pml_write_files = function (x, directory = "solutions") {
  dir.create (directory)
  
  n = length(x)
  for(i in 1:n){
    filename = paste0 ("problem_id_",i,".txt")
    filename = file.path (directory, filename)
    write.table (x[i], file=filename, quote=FALSE, row.names=FALSE, col.names=FALSE)
  }
}

pml_write_files (answers)
```


